---
title: "Assignment 4 - Coordinating Heart Rate"
author: "Riccardo Fusaroli"
date: "November 6, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysing Heart Rate and Respiration data

The goal of this assignment is to first familiarize you with heart rate, and respiration data and their preprocessing. The second part explores how to analyze interpersonal coordination of these signals.

These are the questions you need to be able to answer at the end of the assignment (aka that you need to submit as part of the portfolio)


```{r}
library(pacman)
library(crqa)
library(tidyverse)
library(dplyr)
library(stringr)
p_load(groupdata2)

```

```{r}
############### ON ONE DATASET #########


# READ IN DATA
filename = "CleanData/Study2_G5_T1_Synchronous.csv"
d1 = read.csv (filename, header = T)

# EXTRACT DATA FROM NAME
Study = str_extract(filename, "Study+\\d+") %>%   
      str_extract("\\d+")
Group = str_extract(filename, "G+\\d+") %>%   
      str_extract("\\d+")
Trial = str_extract(filename, "T+\\d+") %>%   
      str_extract("\\d+")
Condition = gsub(".*[_]([^.]+)[.].*", "\\1", filename)
  
#DOWNSAMPLE
p_load(groupdata2)
d1 = d1 %>%
  group(n= 100, method= 'greedy') %>% 
  dplyr::summarise(
    time= mean(time,na.rm=T),
    HR1 = mean(HR1,na.rm=T),
    HR2 = mean(HR2,na.rm=T),
    Resp1 = mean(Resp1,na.rm=T),
    Resp2 = mean(Resp2,na.rm=T))

#ARTIFACTS
removeOuts <- function(ts,threshold){
  ts[ts > (mean(ts,na.rm=T) +
             (threshold*sd(ts,na.rm=T))) |
       ts < (mean(ts,na.rm=T) -
             (threshold*sd(ts,na.rm=T)))] =
  mean(ts,na.rm=T)
  return(ts)}
threshold=2.5
d1$HR1=removeOuts(d1$HR1,threshold) 
d1$HR2=removeOuts(d1$HR2,threshold) 
d1$Resp1=removeOuts(d1$Resp1,threshold) 
d1$Resp2=removeOuts(d1$Resp2,threshold) 


#SCALE
d1$Resp1=scale(d1$Resp1)
d1$Resp2=scale(d1$Resp2)
d1$HR1=scale(d1$HR1)
d1$HR2=scale(d1$HR2)


#GET RQA RESULTS
#Finding the optimal parameters: run on all pairs:
par = list(lgM =  50, steps = seq(1, 6, 1),  radiusspan = 100,  radiussample = 40, normalize = 0,  rescale = 0,  mindiagline = 2,  minvertline = 2,  tw = 0,  whiteline = FALSE,  recpt = FALSE,  fnnpercent = 10,  typeami = "mindip")
ans = optimizeParam(d1$HR1, d1$HR2, par, min.rec = 2, max.rec = 8) 

Results=crqa (d1$HR1, d1$HR2, delay=10, embed=8, radius=4,normalize=0,rescale=0,mindiagline = 2,minvertline = 2)


#Represent the plot:
RP=Results$RP
RP = matrix(as.numeric(RP), nrow = ncol(RP))
cols = c("white","blue4") 
image(RP, xlab = "", ylab = "", col = cols) 
#Explore the lags of coordination: 
Profile=drpdfromts(d1$HR1, d1$HR2, datatype = 'continuous',ws=50,radius=4) 
timecourse = round( seq(-5000,5000,100)/1000, digit = 1) 
maxlag = Profile$maxlag/1000 
profile = Profile$profile*100 
Prof=data.frame(profile) 
ggplot(Prof, aes(timecourse,profile))+geom_line()+ geom_vline(xintercept = timecourse[maxlag], colour='red')



```


1) How do you preprocess heart rate and respiration data? Describe the process. If any data needs to be excluded, list the excluded data and motivate the exclusion.


```{r}
############  ALL DATASETS ###############


####### PREPROCESSING #####
#Get a list of all the filenames
Files0 = list.files("CleanData/", ".csv")
Files0_t = sample(Files0, 6)
#Iteration count
n = 1


for (i in Files0) {
#Put the path together with the filename
  dest = paste("CleanData/", i, sep="")
  #Read it
  d_t <- read.csv(dest, header = T)
  
  #DOWNSAMPLE
  d_t = d_t %>%
    group(n= 100, method= 'greedy') %>% 
    dplyr::summarise(
      time= mean(time,na.rm=T),
      HR1 = mean(HR1,na.rm=T),
      HR2 = mean(HR2,na.rm=T),
      Resp1 = mean(Resp1,na.rm=T),
      Resp2 = mean(Resp2,na.rm=T))

  #ARTIFACTS
  removeOuts <- function(ts,threshold){
    ts[ts > (mean(ts,na.rm=T) +
               (threshold*sd(ts,na.rm=T))) |
         ts < (mean(ts,na.rm=T) -
               (threshold*sd(ts,na.rm=T)))] =
    mean(ts,na.rm=T)
    return(ts)}

  threshold=2.5

  d_t$HR1=removeOuts(d_t$HR1,threshold) 
  d_t$HR2=removeOuts(d_t$HR2,threshold) 
  d_t$Resp1=removeOuts(d_t$Resp1,threshold) 
  d_t$Resp2=removeOuts(d_t$Resp2,threshold) 

  #SCALE
  d_t$Resp1=scale(d_t$Resp1)
  d_t$Resp2=scale(d_t$Resp2)
  d_t$HR1=scale(d_t$HR1)
  d_t$HR2=scale(d_t$HR2)
    
  NewFilename = paste("PrepData/Prep", i, sep="")
  write.csv(d_t, file = NewFilename )
  
  print (n)
  n = n+1
}




```



```{r}

######## DETERMINING THE OPTIMAL PARAMETERS
#Get a list of all the files
Files = list.files("CleanData/", ".csv")
#Get a smaller list for experimentation
Files_t = sample(Files, 6)


#Make lists for values
Delay = NULL
Embed = NULL
Radius = NULL
#Iteration count
n = 1

#Define input to parameter optimization
par = list(lgM =  50, steps = seq(1, 6, 1),  radiusspan = 100,  radiussample = 40, normalize = 0,  rescale = 0,  mindiagline = 2,  minvertline = 2,  tw = 0,  whiteline = FALSE,  recpt = FALSE,  fnnpercent = 10,  typeami = "mindip")

#NOTE: I use the normal, most conservative settings for parameter optimization. But I made the rec limits wider.

#Optimize for each file
for (i in Files_t){
  
  #Put the path together with the filename
  dest = paste("CleanData/", i, sep="")
  #Read it
  d_t <- read.csv(dest, header = T)
  

  #Get optimized parameters
  ans = try(optimizeParam(d_t$HR1, d_t$HR2, par, min.rec= 2, max.rec= 8))
  
  if (length(ans) < 2){
    Delay[n] = NA
    Embed[n] = NA
    Radius[n] = NA
  } else {
    Delay[n] = ans$delay
    Embed[n] = ans$emddim
    Radius[n] = ans$radius}
  
  #loop end
  print (n)
  n = n+1 
  }

DELAY = median(Delay, na.rm = T)
EMBED = median(Embed, na.rm = T)
RADIUS = median(Radius, na.rm = T) + 1.2







# LOOP ABOVE THROUGH REAL PAIRS

#Create lists for results
Study = NULL
Group = NULL
Trial = NULL
Condition = NULL
#CQRA
HR_RR = NULL
HR_DET = NULL
HR_NRLINE = NULL
HR_maxL = NULL
HR_L = NULL
HR_ENTR = NULL
HR_LAM = NULL
HR_TT = NULL
#iteration counter
n = 1



for (i in Files_t){
  ###Read file
  #Put the path together with the filename
  dest = paste("CleanData/", i, sep="")
  #Read it
  d_t <- read.csv(dest, header = T)
  
  #Extract from filename
  Study [n] = str_extract(i, "Study+\\d+") %>%   
        str_extract("\\d+")
  Group [n] = str_extract(i, "G+\\d+") %>%   
        str_extract("\\d+")
  Trial [n] = str_extract(i, "T+\\d+") %>%   
        str_extract("\\d+")
  Condition [n] = gsub(".*[_]([^.]+)[.].*", "\\1", i)
  
  #Extract descriptive statistics
  P_Mean[n] = mean(d_t$f0)
  P_SD[n] = sd(d_t$f0)
  P_Range[n] = range(d_t$f0)[2] - range(d_t$f0)[1]
  P_Median[n] = median(d_t$f0)
  P_IQR[n] = IQR(d_t$f0)
  P_MeanAbDif[n] = mean(abs(d_t$f0-mean(d_t$f0)))
  P_CovofVar[n] = sd(d_t$f0) / mean(d_t$f0)
  
  #Do CRQA analysis
    crqa_t = try(crqa (d_t, d_t, delay=DELAY, embed=EMBED, radius=RADIUS, normalize=0, rescale=0, mindiagline = 2, minvertline = 2)) 
  
  
  if (is.na(Delay[n]) | length(crqa_t) < 2 ){
    P_RR[n] = NA
    P_DET[n] = NA
    P_NRLINE[n] = NA
    P_maxL[n] = NA
    P_L[n] = NA
    P_ENTR[n] = NA
    P_LAM[n] = NA
    P_TT[n] = NA
  } else {
    P_RR[n] = crqa_t$RR
    P_DET[n] = crqa_t$DET
    P_NRLINE[n] = crqa_t$NRLINE
    P_maxL[n] = crqa_t$maxL
    P_L[n] = crqa_t$L
    P_ENTR[n] = crqa_t$ENTR
    P_LAM[n] = crqa_t$LAM
    P_TT[n] = crqa_t$TT}
  
  #Loop end
  print(n)
  n = n+1}






# READ IN DATA
filename = "CleanData/Study2_G5_T1_Synchronous.csv"
d1 = read.csv (filename, header = T)

# EXTRACT DATA FROM NAME
Study = str_extract(filename, "Study+\\d+") %>%   
      str_extract("\\d+")
Group = str_extract(filename, "G+\\d+") %>%   
      str_extract("\\d+")
Trial = str_extract(filename, "T+\\d+") %>%   
      str_extract("\\d+")
Condition = gsub(".*[_]([^.]+)[.].*", "\\1", filename)
  
#DOWNSAMPLE
p_load(groupdata2)
d1 = d1 %>%
  group(n= 100, method= 'greedy') %>% 
  dplyr::summarise(
    time= mean(time,na.rm=T),
    HR1 = mean(HR1,na.rm=T),
    HR2 = mean(HR2,na.rm=T),
    Resp1 = mean(Resp1,na.rm=T),
    Resp2 = mean(Resp2,na.rm=T))

#ARTIFACTS
removeOuts <- function(ts,threshold){
  ts[ts > (mean(ts,na.rm=T) +
             (threshold*sd(ts,na.rm=T))) |
       ts < (mean(ts,na.rm=T) -
             (threshold*sd(ts,na.rm=T)))] =
  mean(ts,na.rm=T)
  return(ts)}

threshold=2.5

d1$HR1=removeOuts(d1$HR1,threshold) 
d1$HR2=removeOuts(d1$HR2,threshold) 
d1$Resp1=removeOuts(d1$Resp1,threshold) 
d1$Resp2=removeOuts(d1$Resp2,threshold) 


#SCALE
d1$Resp1=scale(d1$Resp1)
d1$Resp2=scale(d1$Resp2)
d1$HR1=scale(d1$HR1)
d1$HR2=scale(d1$HR2)


#GET RQA RESULTS
#Finding the optimal parameters: run on all pairs:
par = list(lgM =  50, steps = seq(1, 6, 1),  radiusspan = 100,  radiussample = 40, normalize = 0,  rescale = 0,  mindiagline = 2,  minvertline = 2,  tw = 0,  whiteline = FALSE,  recpt = FALSE,  fnnpercent = 10,  typeami = "mindip")
ans = optimizeParam(d1$HR1, d1$HR2, par, min.rec = 2, max.rec = 8) 

Results=crqa (d1$HR1, d1$HR2, delay=10, embed=8, radius=4,normalize=0,rescale=0,mindiagline = 2,minvertline = 2)


#Represent the plot:
RP=Results$RP
RP = matrix(as.numeric(RP), nrow = ncol(RP))
cols = c("white","blue4") 
image(RP, xlab = "", ylab = "", col = cols) 
#Explore the lags of coordination: 
Profile=drpdfromts(d1$HR1, d1$HR2, datatype = 'continuous',ws=50,radius=4) 
timecourse = round( seq(-5000,5000,100)/1000, digit = 1) 
maxlag = Profile$maxlag/1000 
profile = Profile$profile*100 
Prof=data.frame(profile) 
ggplot(Prof, aes(timecourse,profile))+geom_line()+ geom_vline(xintercept = timecourse[maxlag], colour='red')



#CREATE SHUFFLED PAIRS - LOOP CRQA ON THEM




#CREATE SURROGATE PAIRS - LOOP CRQA

```


2) Do you observe interpersonal coordination in heart rate and respiration? Describe your control baseline, the method used to quantify coordination, and the statistical models used to infer whether coordination was higher than in the baseline. Report the results of the models.


```{r}






```


3) Do you observe differences in coordination between conditions? Report the models and results.


```{r}
#####CONDITIONS


#If do interaction, will take away the stuff from doing the same activity
HR_RR ~ Condition*Type


```

4) Is respiration coordination a likely driver of heart rate coordination? Describe how you would test for it. Bonus points if you actually run the tests and report methods and results.


```{r}



```


N.B. to give you a bit more data I included data from last year (Study1) and from your class (Study2). Note that synchronouns and turn-taking are the same across both studies, but the third condition is different: last year it was self-paced joint reading; this year it was the tv-series conversation. So you might want to exclude the self-paced reading (but, up to you!)

## Step by step suggestions to solve the assignment

### Exploring physiological signals

- Choose one pair (one pair, three conditions)
- Load the logs
- Produce a plot of the participants' respiration signal and a different one of the participants' HR signal (for inspecting whether the data is usable)
  N.B: remember the slides: artifacts, downsampling, scaling.
  N.B. The gridExtra::grid.arrange() function allows you to display the plots side by side. E.g. grid.arrange(plot1, plot2, plot3, ncol=3)
- Can you eye-ball which condition if any displays more physiological coordination?

- Run crqa on heart rate and respiration data (find parameters, run crqa)
- Does this tell you more than just eyeballing the plots?

### Systematically pre-process the data
- Loop through all the files (either with a loop or with a function), check which files should be excluded, if any, and save the pre-processed time-series. Tip: plot and visually inspect the data to figure out which should be excluded.
- Run crqa on all the pre-processed time-series and save the output (don't forget to add columns with study, group, condition and trial). Tip: remember to first assess optimal parameters (dimensions, delay, radius) across all timeseries. Tip: it will often fail, just take whatever parameters you get, select optimal across timeseries parameters and run crqa on all timeseries with those. Tip: double check the rr. When I ran the loop, I got very low rr, so I adjusted the radius until the average of rr across all pairs was approx. 4%.

### Creating controls: shuffled controls
 - loop through all pairs and conditions
 - shuffle the timeseries (take a timeseries and rearrange its values in a random order). Tip check the sample() function
 - run crqa and save the output. NB. which delay, embed, radius parameters should you use?
 - statistically compare the crqa indexes in real and shuffled pairs
 
### TRICKY! Creating controls: surrogate pair controls
 - Per each real pair, identify at least one surrogate pair (matching one of the participants, with somebody doing the same task, but in a different pair). Tip: Celine will share a commented script
 - Run crqa on all the surrogate pairs and save the output. NB. which delay, embed, radius parameters should you use?
 - Test whether crqa shows a difference between real and surrogate pairs

### Testing effects of conditions
 - make a (probably underpowered) mixed model testing effects of the different conditions on heart rate and respiration coordination
 - N.B: would it make sense to include surrogate pairs? and if so how? what would that tell you?

### Effects of respiration coordination on heart rate coordination
 - describe how you would test those.
 - Optional: run the models and report them